\documentclass[14pt,aspectratio=169]{beamer}
\usetheme{iiasa}

\usepackage{ifthen}
\ifthenelse{\equal{\detokenize{notes}}{\jobname}}{%
\setbeameroption{show only notes}%
}{
%
}

\usepackage[
  maxnames = 1,
  style = authoryear,
  giveninits,
  terseinits,
  maxcitenames = 3,
  ]{biblatex}
\addbibresource{all.bib}
% Small font in bibliography
\renewcommand*\bibfont{\small}
\addbibresource{all.bib}

\usepackage{minted}
\usepackage{pdfpages}

\usepackage{tikz}
\usetikzlibrary{calc}

\title{Data for modeling}
\subtitle{User roundtable}
\institute{Energy, Climate, and Environment (ECE) Program \\
  International Institute for Applied Systems Analysis (IIASA)}

\date{\texorpdfstring{Message Community Meeting\\
  2023-05-25}%
  {2023-05-25}}

\author{\texorpdfstring{Dr. Paul Natsuo Kishimoto \scriptsize\newline
  \href{mailto:paul.kishimoto@iiasa.ac.at}%
       {\ttfamily <paul.kishimoto@iiasa.ac.at>}}%
  {Dr. Paul Natsuo Kishimoto <paul.kishimoto@iiasa.ac.at>}}

\begin{document}

\maketitle

\begin{frame}
  \frametitle{Roundtable, interactive discussion}

  Each of us works with model data:

  \begin{itemize}
    \item Preparing input data to build models:
          \begin{itemize}
            \item Parameter values.
            \item Reference for calibration.
          \end{itemize}
    \item Reporting model results.
    \item General research/analysis based on model results \emph{combined with} other data.
  \end{itemize}

  \bigskip
  \structure{\bfseries Task:} think about a particular kind of data or data work that you do in your own research.
  Make some notes on your laptop, mobile, or paper.

\end{frame}


\begin{frame}
  \frametitle{Data work can be costly}

  Costs: our (colleagues') time, money, direct monetary cost.

  \bigskip
  Some data-related work that can take a long time:

  \begin{columns}
    \column[t]{0.5\textwidth}
    \begin{itemize}
      \item Find sources for particular measures.
      \item Check quality (many criteria) of source data.
      \item Relabel data.
      \item Add/remove dimensions.
    \end{itemize}

    \column[t]{0.5\textwidth}
    \begin{itemize}
      \item (Dis)aggregate along 1 or more dimensions.
      \item Transform or derive a related measure.
      \item Check processed data.
      \item Visualize the data.
      \item Document the steps taken.
    \end{itemize}
  \end{columns}

\end{frame}

\begin{frame}
  \frametitle{Share with colleagues}
  \framesubtitle{Alternate in-room and on-line participants}

  What is the particular data work that has been costly for you?

  Examples:
  \begin{itemize}
    \item Work with input data from a specific source.
    \item Prepare model output data for a specific purpose.
  \end{itemize}

  \pause
  \bigskip
  What tasks or steps in working with these data were difficult?

  Why?

  \pause
  \bigskip
  What tools or techniques have you tried to help reduce these costs?

  Did they help? How, or why not?
\end{frame}

\end{document}
